---
title: "Brain charts across sites"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brain charts across sites}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  dpi = 96
)
```

The standard normative-modelling workflow in neuroimaging is: pool
controls from several scanners, estimate the reference distribution of
each regional measure as a function of age and sex with a site effect,
score every reference subject out of sample, and then score new
participants, often from a scanner that contributed no controls. This
vignette runs that workflow end to end on a simulated panel and
compares the three ways of scoring a new site.

```{r setup}
library(referent)
```

## A multi-site reference panel

Five sites, with site-specific level shifts and one site with a wider
spread. The outcome `y` plays the role of a regional volume; the three
markers are further regions that share a latent factor with `y`.

```{r data}
ref <- ref_simulate(
  800, sites = 5,
  site_shift = c(0, 0.8, -0.5, 0.3, 0),
  site_log_scale = c(0, 0, 0, 0, 0.2),
  seed = 1
)
ref$participant_id <- sprintf("R%04d", seq_len(nrow(ref)))
regions <- c("y", "marker_01", "marker_02", "marker_03")
table(ref$site)
```

The site effect enters the location as a random effect, `s(site, bs =
"re")`. Unlike a fixed site factor this shrinks small sites toward the
pooled curve and, more importantly, gives a well-defined prediction for
a level that was not in the reference: the random effect is set to zero
and the population curve is used. `referent` reports those rows with
`support == "new_group"`.

```{r fit}
spec <- ref_spec(
  family = ref_gaussian(),
  location = ~ s(age, k = 6) + sex + s(site, bs = "re"),
  scale = ~1
)
fit <- ref_fit(spec, data = ref, outcomes = all_of(regions), id = participant_id)
fit
tidy(fit)[, c("outcome", "status", "n", "edf_s(age)", "edf_s(site)")]
```

```{r centiles}
autoplot(fit, type = "centiles", outcome = "y", by = sex)
```

## Out-of-fold reference scores

Scoring the reference rows with the model that was fit to them
understates their deviations. `ref_crossfit()` gives every reference
row a score from a fold that excluded it, stratified by site so that
each fold's model still sees every scanner. These scores are the honest
reference for anything downstream: calibration checks, the joint model,
and the expected exceedance rate.

```{r crossfit}
cf <- ref_crossfit(spec, data = ref, outcomes = all_of(regions), folds = 5,
                    strata = site, id = participant_id)
oof <- split(cf, cf$.outcome)
oof_calibration <- data.frame(
  region = names(oof),
  n = sapply(oof, nrow),
  n_scored = sapply(oof, function(s) sum(is.finite(s$z))),
  mean_z = round(sapply(oof, function(s) mean(s$z, na.rm = TRUE)), 3),
  var_z = round(sapply(oof, function(s) var(s$z, na.rm = TRUE)), 3),
  cover_95 = round(sapply(oof, function(s) mean(abs(s$z) < qnorm(0.975), na.rm = TRUE)), 3),
  crps = round(sapply(oof, function(s) mean(s$crps, na.rm = TRUE)), 3),
  row.names = NULL
)
oof_calibration
```

A few rows have no out-of-fold `z`: their age lies outside the range of
the fold that scored them, so `support` is `"out"` and `z` is masked.
That is the extrapolation rule doing its job, not a failure.

## Scoring a new site

A sixth scanner arrives with 200 participants. Its level differs from
every reference site. Thirty of its participants are known controls; the
rest are the people we want to score.

```{r newsite}
new_site <- ref_simulate(
  1500, sites = 6,
  site_shift = c(0, 0.8, -0.5, 0.3, 0, 1.0),
  site_log_scale = c(0, 0, 0, 0, 0.2, 0),
  seed = 2
)
new_site <- new_site[new_site$site == "F", ][1:200, ]
new_site$participant_id <- sprintf("F%04d", seq_len(nrow(new_site)))
controls <- new_site[1:30, ]
held_out <- new_site[31:200, ]
table(ref_support(fit, held_out)$support)
```

There are three options, and they answer different questions.

1. **Score against the population curve.** Use the fit as it is. The
   random effect for site F is zero, so the scores are relative to the
   average reference site. Any level difference of the scanner becomes a
   shift in everyone's `z`.
2. **Adapt.** `ref_adapt()` freezes the shared trajectory and estimates
   a shrunk location (and optionally scale) offset for site F from the
   30 controls. This is the right tool when the scanner moves the
   measurements but the age trend is shared.
3. **Calibrate by site.** `ref_calibrate(by = site)` re-maps the PIT
   values of site F through their empirical distribution on the 30
   controls. It fixes probabilities rather than parameters and needs
   more local data to do so without noise.
4. **Refit** with the 30 controls appended and site F as a sixth level.
   This is the most flexible but the least stable with few controls, and
   it changes the reference for everyone else.

```{r strategies}
adapted <- ref_adapt(fit, data = controls, by = site,
                      parameters = c("location", "scale"))
calibrated <- ref_calibrate(fit, data = controls, by = site)
refit <- ref_fit(spec, data = rbind(ref, controls), outcomes = all_of(regions),
                  id = participant_id)
adapted$adaptation
```

Each strategy is judged on the 170 held-out site-F rows. Because site F
was simulated from the same generator as the reference, a correct
strategy must return calibrated scores there: `mean_z` near 0, `var_z`
near 1, 95% coverage near 0.95. Thirty controls bound what any local
strategy can achieve: the standard error of a 30-person mean `z` is
about 0.18, and the controls themselves sit where chance put them
relative to the rest of site F.

```{r control-anchor}
ctrl_scores <- predict(fit, newdata = controls, type = "scores", outcomes = "y")
c(controls_mean_z = round(mean(ctrl_scores$z), 2),
  held_out_mean_z = round(mean(predict(fit, newdata = held_out, type = "scores",
                                       outcomes = "y")$z), 2))
```

```{r compare}
strategies <- list(population = fit, adapted = adapted, calibrated = calibrated, refit = refit)
comparison <- do.call(rbind, lapply(names(strategies), function(nm) {
  a <- ref_assess(strategies[[nm]], newdata = held_out)
  m <- a$marginal[a$marginal$.outcome == "y", ]
  data.frame(
    strategy = nm,
    mean_z = round(m$mean_z, 2),
    var_z = round(m$var_z, 2),
    cover_95 = round(m$cover_95, 3),
    log_score = round(a$overall$mean_log_score[a$overall$.outcome == "y"], 3)
  )
}))
comparison
```

The population-curve scores carry the scanner's level shift into
`mean_z` and pay for it in log score. The three local strategies remove
most of the shift, and what remains is the difference between the 30
controls and the other 170 participants, which none of them can know.
They differ in how they spend the 30 observations. Adaptation estimates
two numbers (a shrunk location and log-scale offset) on the family's own
likelihood and leaves the age curve alone, so it is the most stable.
Calibration by site re-maps probabilities through a 30-knot rank map;
its centre is as good as the controls' mean, but its spread and tails
are noisy at this sample size. The refit lets the controls move the
shared smooth and the random-effect variance as well as the site level,
and it changes the reference for every other site. With 20-50 local
controls, adaptation is the default recommendation; calibration earns
its place when the distributional family, not the level, is wrong at
the new site, and then needs a few hundred local rows.

```{r adaptplot}
autoplot(adapted, type = "adaptation")
```

## Joint deviation across regions

Per-region Z-scores are correlated; a participant who is low on every
region is more unusual than four independent `z = -1.5` values suggest.
`ref_joint()` fits a Gaussian copula on the out-of-fold reference
scores, with a shrinkage correlation estimate, and returns a joint
centile calibrated through the reference distribution of the
Mahalanobis statistic. Missing regions are handled by the observed
submatrix.

```{r joint}
joint <- ref_joint(cf)
joint
new_scores <- predict(adapted, newdata = held_out, type = "scores")
joint_scores <- predict(joint, new_scores)
head(joint_scores)
mean(joint_scores$joint_centile > 0.95, na.rm = TRUE)
```

## Flags, and how many to expect

`ref_flag()` marks `|z| > 2` per region and adds FDR-adjusted two-sided
tail probabilities across all region-by-person tests. Under calibration
about 4.6% of rows exceed the threshold by chance; the attribute
`expected_exceedances` is that count.

```{r flag}
flags <- ref_flag(new_scores, threshold = 2)
c(observed = attr(flags, "observed_exceedances"),
  expected = round(attr(flags, "expected_exceedances"), 1))
sum(flags$fdr < 0.05, na.rm = TRUE)
autoplot(new_scores[new_scores$.row <= 15, ], type = "heatmap")
```

## What to report

* The reference: `ref_freeze(fit)` prints a model card with the
  family, formulas, covariate ranges, site levels, and software
  versions. Ship the frozen bundle, not the training data.
* Out-of-fold calibration of the reference (the `oof_calibration` table),
  which is the evidence that the scores mean what they claim.
* For each new site: how it was scored (population, adapted,
  calibrated, or refit), the number of local controls, the estimated
  offsets with their standard errors, and the held-out calibration if
  any controls were held back.
* The expected and observed exceedance counts, and the joint centiles,
  rather than a list of flagged people.

```{r card}
ref_freeze(adapted, criteria = "healthy controls, 20-80 years", units = "mm^3")
```
