---
title: "Brain charts across sites"
description: "Build multi-site brain charts, transport them to a new scanner, and report calibrated deviation scores."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brain charts across sites}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  dpi = 96
)
```

The standard normative-modelling workflow in neuroimaging is: pool
controls from several scanners, estimate the reference distribution of
each regional measure as a function of age and sex with a site effect,
score every reference subject out of sample, and then score new
participants, often from a scanner that contributed no controls. This
vignette runs that workflow end to end on a simulated panel and
compares the three ways of scoring a new site.

```{r setup}
library(referent)
```

# A multi-site reference panel

Five sites, with site-specific level shifts and one site with a wider
spread. The outcome `y` plays the role of a regional volume; the three
markers are further regions that share a latent factor with `y`.

```{r data}
ref <- ref_simulate(
  800, sites = 5,
  site_shift = c(0, 0.8, -0.5, 0.3, 0),
  site_log_scale = c(0, 0, 0, 0, 0.2),
  seed = 1
)
ref$participant_id <- sprintf("R%04d", seq_len(nrow(ref)))
regions <- c("y", "marker_01", "marker_02", "marker_03")
table(ref$site)
```

The site effect enters the location as a random effect, `s(site, bs =
"re")`. Unlike a fixed site factor this shrinks small sites toward the
pooled curve and, more importantly, gives a well-defined prediction for
a level that was not in the reference: the random effect is set to zero
and the population curve is used. `referent` reports those rows with
`support == "new_group"`.

```{r fit}
spec <- ref_spec(
  family = ref_gaussian(),
  location = ~ s(age, k = 6) + sex + s(site, bs = "re"),
  scale = ~1
)
fit <- ref_fit(spec, data = ref, outcomes = all_of(regions), id = participant_id)
fit
tidy(fit)[, c("outcome", "status", "n", "edf_s(age)", "edf_s(site)")]
```

```{r centiles, fig.alt="Reference centile bands for one simulated brain region across age, split by sex."}
autoplot(fit, type = "centiles", outcome = "y", by = sex)
```

# Scaling to many outcomes

A regional panel has tens of outcomes; a vertex-wise or voxel-wise chart has
tens of thousands. `ref_fit()` fits each outcome independently, so the
problem is embarrassingly parallel, and the package has three levers for it.

**Parallel fits in one session.** When `future.apply` is installed and a
non-sequential plan is active, outcomes are fitted across workers. Nothing
else changes: the same call, the same object back.

```{r parallel, eval = FALSE}
future::plan(future::multisession, workers = 8)
fit_wide <- ref_fit(spec, data = wide, outcomes = starts_with("vertex_"))
future::plan(future::sequential)
```

**Memory per outcome, and why to freeze.** Each fitted outcome carries its
`mgcv` model, including the model frame, fitted values, and residuals; those
are what make a fit large. `ref_freeze()` keeps only what prediction needs
(coefficients, their covariance, the smooth constructions, factor levels)
and reproduces the fit's predictions exactly. On 20 outcomes at n = 5000,
the fit is about 14 MB and the bundle about 3 MB, a factor of 4.6; the
saving grows with n because the bundle's size does not depend on it.

```{r freeze-size}
wide <- ref_simulate(5000, seed = 1)
set.seed(2)
for (j in 1:20) wide[[sprintf("vertex_%02d", j)]] <- wide$y + rnorm(5000)
wide_spec <- ref_spec(ref_gaussian(), ~ s(age, k = 6) + sex, scale = ~1)
fit_wide <- ref_fit(wide_spec, wide, outcomes = starts_with("vertex_"))
bundle_wide <- ref_freeze(fit_wide)
c(fit = format(object.size(fit_wide), units = "MB"),
  bundle = format(object.size(bundle_wide), units = "MB"))
```

**Blocks as independent jobs.** For a matrix too wide to hold in one
session, split the outcome columns into blocks, fit each block as its own
job, and combine the score tables afterwards. Outcomes share no state in
`ref_fit()`, so the combined result is identical to one big fit. The demo
below fits 12 outcomes as three blocks of four and checks that claim.

```{r blocks}
vertices <- sprintf("vertex_%02d", 1:12)
target <- ref_simulate(50, seed = 3)
for (v in vertices) target[[v]] <- target$y + rnorm(50)
single <- ref_fit(wide_spec, wide, outcomes = all_of(vertices))
blocks <- split(vertices, rep(1:3, each = 4))
block_fits <- lapply(blocks, function(b) ref_fit(wide_spec, wide, outcomes = all_of(b)))

z_single <- predict(single, target, uncertainty = "conditional")
z_blocks <- vctrs::vec_rbind(!!!lapply(block_fits, predict, newdata = target,
                                       uncertainty = "conditional"))
identical(z_single$z, z_blocks$z)
```

The same holds for `augment()`: the wide per-observation frames from each
block share the covariate columns and can be joined on `.row`. On a cluster,
the block is the array index. A `fit_block.R` that reads its block of
columns, fits, freezes, and saves the bundle is the whole job:

```{r slurm, eval = FALSE}
# fit_block.R, run as: Rscript fit_block.R $SLURM_ARRAY_TASK_ID
block <- as.integer(commandArgs(trailingOnly = TRUE)[[1]])
columns <- readRDS("outcome_blocks.rds")[[block]]   # list of column-name vectors
wide <- arrow::read_parquet("reference.parquet", col_select = c("age", "sex", "site", columns))
fit <- ref_fit(spec, wide, outcomes = all_of(columns))
ref_write(ref_freeze(fit), sprintf("bundles/block_%04d.rds", block))
```

```
#SBATCH --array=1-100
#SBATCH --cpus-per-task=1 --mem=4G
Rscript fit_block.R $SLURM_ARRAY_TASK_ID
```

Scoring a new cohort is then `vctrs::vec_rbind()` over
`predict(ref_read(path, trusted = TRUE), newdata)` for each trusted bundle.
Because bundles are small,
the 100 of them for 100k vertices are a few hundred megabytes, and scoring
reads them one at a time.

# Out-of-fold reference scores

Scoring the reference rows with the model that was fit to them
understates their deviations. `ref_crossfit()` gives every reference
row a score from a fold that excluded it, stratified by site so that
each fold's model still sees every scanner. These scores are the honest
reference for anything downstream: calibration checks, the joint model,
and the expected exceedance rate.

```{r crossfit}
cf <- ref_crossfit(spec, data = ref, outcomes = all_of(regions), folds = 5,
                    strata = site, id = participant_id)
oof <- split(cf, cf$.outcome)
oof_calibration <- data.frame(
  region = names(oof),
  n = sapply(oof, nrow),
  n_scored = sapply(oof, function(s) sum(is.finite(s$z))),
  mean_z = round(sapply(oof, function(s) mean(s$z, na.rm = TRUE)), 3),
  var_z = round(sapply(oof, function(s) var(s$z, na.rm = TRUE)), 3),
  cover_95 = round(sapply(oof, function(s) mean(abs(s$z) < qnorm(0.975), na.rm = TRUE)), 3),
  crps = round(sapply(oof, function(s) mean(s$crps, na.rm = TRUE)), 3),
  row.names = NULL
)
oof_calibration
```

A few rows have no out-of-fold `z`: their age lies outside the range of
the fold that scored them, so `support` is `"out"` and `z` is masked.
That is the extrapolation rule doing its job, not a failure.

# Scoring a new site

A sixth scanner arrives with 200 participants. Its level differs from
every reference site. Thirty of its participants are known controls; the
rest are the people we want to score.

```{r newsite}
new_site <- ref_simulate(
  1500, sites = 6,
  site_shift = c(0, 0.8, -0.5, 0.3, 0, 1.0),
  site_log_scale = c(0, 0, 0, 0, 0.2, 0),
  seed = 2
)
new_site <- new_site[new_site$site == "F", ][1:200, ]
new_site$participant_id <- sprintf("F%04d", seq_len(nrow(new_site)))
controls <- new_site[1:30, ]
held_out <- new_site[31:200, ]
table(ref_support(fit, held_out)$support)
```

There are four options, and they answer different questions.

1. **Score against the population curve.** Use the fit as it is. The
   random effect for site F is zero, so the scores are relative to the
   average reference site. Any level difference of the scanner becomes a
   shift in everyone's `z`.
2. **Adapt.** `ref_adapt()` freezes the shared trajectory and estimates
   a shrunk location (and optionally scale) offset for site F from the
   30 controls. This is the right tool when the scanner moves the
   measurements but the age trend is shared.
3. **Calibrate by site.** `ref_calibrate(by = site)` re-maps the PIT
   values of site F through their empirical distribution on the 30
   controls. It fixes probabilities rather than parameters and needs
   more local data to do so without noise.
4. **Refit** with the 30 controls appended and site F as a sixth level.
   This is the most flexible but the least stable with few controls, and
   it changes the reference for everyone else.

```{r strategies}
adapted <- ref_adapt(fit, data = controls, by = site,
                      parameters = c("location", "scale"))
calibrated <- ref_calibrate(fit, data = controls, by = site)
refit <- ref_fit(spec, data = rbind(ref, controls), outcomes = all_of(regions),
                  id = participant_id)
adapted$adaptation
```

The first three objects still record site F as `support == "new_group"`
because the original reference contained no site-F rows. Adaptation or
calibration estimates a transport map; it does not rewrite that provenance.
The calls below therefore set `allow_extrapolation = TRUE` deliberately after
estimating the map from known controls. Without that explicit opt-in,
`referent` masks the scores.

Each strategy is judged on the 170 held-out site-F rows. Because site F
was simulated from the same generator as the reference, a correct
strategy must return calibrated scores there: `mean_z` near 0, `var_z`
near 1, 95% coverage near 0.95. Thirty controls bound what any local
strategy can achieve: the standard error of a 30-person mean `z` is
about 0.18, and the controls themselves sit where chance put them
relative to the rest of site F.

```{r control-anchor}
ctrl_scores <- predict(fit, newdata = controls, type = "scores", outcomes = "y",
                       allow_extrapolation = TRUE)
c(controls_mean_z = round(mean(ctrl_scores$z), 2),
  held_out_mean_z = round(mean(predict(fit, newdata = held_out, type = "scores",
                                       outcomes = "y",
                                       allow_extrapolation = TRUE)$z), 2))
```

```{r compare}
strategies <- list(population = fit, adapted = adapted, calibrated = calibrated, refit = refit)
strategy_scores <- lapply(strategies, function(model) {
  predict(model, newdata = held_out, type = "scores", outcomes = "y",
          allow_extrapolation = TRUE)
})
comparison <- do.call(rbind, lapply(names(strategy_scores), function(nm) {
  sc <- strategy_scores[[nm]]
  ok <- is.finite(sc$z) & is.finite(sc$log_density)
  data.frame(
    strategy = nm,
    mean_z = round(mean(sc$z[ok]), 2),
    var_z = round(var(sc$z[ok]), 2),
    cover_95 = round(mean(sc$centile[ok] > 0.025 & sc$centile[ok] < 0.975), 3),
    log_score = round(mean(sc$log_density[ok]), 3)
  )
}))
comparison
```

```{r compare-checks, include = FALSE}
stopifnot(
  all(vapply(strategy_scores, function(sc) all(is.finite(sc$z)), logical(1))),
  all(is.finite(comparison$mean_z)),
  all(is.finite(comparison$var_z) & comparison$var_z > 0),
  all(comparison$cover_95 > 0 & comparison$cover_95 <= 1)
)
```

The population-curve scores carry the scanner's level shift into
`mean_z` and pay for it in log score. The three local strategies remove
most of the shift, and what remains is the difference between the 30
controls and the other 170 participants, which none of them can know.
They differ in how they spend the 30 observations. Adaptation estimates
two numbers (a shrunk location and log-scale offset) on the family's own
likelihood and leaves the age curve alone, so it is the most stable.
Calibration by site re-maps probabilities through a four-parameter
sinh-arcsinh map on the normal scores; its centre is as good as the
controls' mean, but its spread and tail parameters are noisy at this sample
size. The refit lets the controls move the
shared smooth and the random-effect variance as well as the site level,
and it changes the reference for every other site. In this simulation with 30
controls, adaptation spends the local information on two parameters and is
more stable than the rank map or refit. That is an illustration, not a general
sample-size rule: choose and validate a transport strategy with held-out local
controls whenever the design permits it.

```{r adaptplot, fig.alt="Estimated location and scale offsets for the transported site with uncertainty intervals."}
autoplot(adapted, type = "adaptation")
```

# Joint deviation across regions

Per-region Z-scores are correlated; a participant who is low on every
region is more unusual than four independent `z = -1.5` values suggest.
`ref_joint()` fits a Gaussian copula on the out-of-fold reference
scores, with a shrinkage correlation estimate, and returns a joint
centile calibrated through the reference distribution of the
Mahalanobis statistic. Missing regions are handled by the observed
submatrix.

```{r joint}
joint <- ref_joint(cf)
joint
new_scores <- predict(adapted, newdata = held_out, type = "scores",
                      allow_extrapolation = TRUE)
joint_scores <- predict(joint, new_scores)
head(joint_scores)
mean(joint_scores$joint_centile > 0.95, na.rm = TRUE)
```

```{r joint-checks, include = FALSE}
stopifnot(
  all(joint_scores$n_observed > 0),
  all(is.finite(joint_scores$joint_centile)),
  all(joint_scores$joint_centile > 0 & joint_scores$joint_centile < 1)
)
```

# Flags, and how many to expect

`ref_flag()` marks `|z| > 2` per region and adds FDR-adjusted two-sided
tail probabilities across all region-by-person tests. Under calibration
about 4.6% of rows exceed the threshold by chance; the attribute
`expected_exceedances` is that count.

```{r flag}
flags <- ref_flag(new_scores, threshold = 2)
c(observed = attr(flags, "observed_exceedances"),
  expected = round(attr(flags, "expected_exceedances"), 1))
sum(flags$fdr < 0.05, na.rm = TRUE)
```

```{r heatmap, fig.alt="Deviation-score heatmap for fifteen held-out participants across four correlated simulated brain regions."}
autoplot(new_scores[new_scores$.row <= 15, ], type = "heatmap")
```

```{r flag-checks, include = FALSE}
stopifnot(
  attr(flags, "expected_exceedances") > 0,
  all(is.finite(flags$tail_prob))
)
```

# What to report

* The reference: `ref_freeze(fit)` prints a model card with the
  family, formulas, covariate ranges, site levels, and software
  versions. Ship the frozen bundle, not the training data.
* Out-of-fold calibration of the reference (the `oof_calibration` table),
  which is the evidence that the scores mean what they claim.
* For each new site: how it was scored (population, adapted,
  calibrated, or refit), the number of local controls, the estimated
  offsets with their standard errors, and the held-out calibration if
  any controls were held back.
* The expected and observed exceedance counts, and the joint centiles,
  rather than a list of flagged people.

```{r card}
ref_freeze(adapted, criteria = "healthy controls, 20-80 years", units = "mm^3")
```

# Ship and reuse a reference

A reference is useful to other sites only if it can be applied without the
training data. `ref_freeze()` produces that object; `ref_write()` persists it.

```{r ship}
bundle <- ref_freeze(fit, criteria = "healthy controls, 20-80 years", units = "mm^3")
bundle_path <- file.path(tempdir(), "reference-v1.rds")
ref_write(bundle, bundle_path)
file.size(bundle_path)
```

The bundle contains, per outcome, the coefficients, their covariance `Vp`,
the smooth bases, the terms and factor levels (`xlevels`), plus the support
reference (covariate ranges and factor levels), the calibration and
adaptation slots if any were estimated, the reference-sample baseline used by
`standardized_log_score`, and the model card. It excludes the training rows,
the model frame, fitted values, and residuals. No individual observation can
be recovered from it, which makes it suitable for privacy-preserving sharing.
That privacy property does not make the RDS container safe to open: RDS is
executable R serialization, so accept bundles only from an authenticated,
trusted source and verify their provenance through an independent channel.

A receiving site needs only the file and the package. Everything that applies
to a `ref_fit` applies to the bundle: scoring, support checks, adaptation, and
calibration.

```{r reuse}
reference <- ref_read(bundle_path, trusted = TRUE)
reference
local_scores <- predict(reference, newdata = held_out, outcomes = "y",
                        allow_extrapolation = TRUE)
table(ref_support(reference, held_out)$support)
local_adapted <- ref_adapt(reference, data = controls, by = site)
local_calibrated <- ref_calibrate(reference, data = controls, by = site)
local_adapted$adaptation
```

The bundle's predictions are the fit's predictions, not an approximation:

```{r bundle-equality}
identical(
  predict(fit, held_out, outcomes = "y", uncertainty = "conditional",
          allow_extrapolation = TRUE)$z,
  predict(reference, held_out, outcomes = "y", uncertainty = "conditional",
          allow_extrapolation = TRUE)$z
)
```

**Versioning.** The printed card records the package, `mgcv`, and R versions
that produced the bundle, and `bundle$data_hash` is a fingerprint of the
training data: its row count, the covariate and outcome names, and a
per-column sum. It distinguishes references fitted to different data without
storing any of it; add `rlang::hash(ref)` to the bundle yourself if you need
a cryptographic digest. Name files by reference version
(`reference-v1.rds`), not by date, and treat a change in the reference
cohort, the specification, or the package major version as a new reference:
scores from different references are not comparable, and the fingerprint is
how a reader can tell two bundles apart.

```{r provenance}
bundle$versions
bundle$data_hash
```

Continue with [Longitudinal change](https://bbuchsbaum.github.io/referent/articles/longitudinal-change/)
(`vignette("longitudinal-change", package = "referent")`), or use
[Troubleshooting reference models](https://bbuchsbaum.github.io/referent/articles/troubleshooting/) when transported
scores remain masked or unstable.
