---
title: "Getting started with referent"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with referent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  dpi = 96
)
```

A reference model answers one question: given a person's covariates, what
distribution of outcome values would the reference population show? Every
other quantity in `referent` is read off that conditional distribution.
A centile is its CDF at the observed value, a Z-score is the normal
quantile of that centile, and a tail probability is how much of the
distribution lies beyond the observation. None of these is a clinical
label.

This vignette walks the standard path on a synthetic growth-like measure:
fit a reference, choose among candidate distributions out of sample,
check calibration on held-out people, and place new observations.

```{r setup}
library(referent)
```

## Reference data

`norm_simulate()` produces data with a known generative structure. The
outcome `y` has a location that bends with age and differs by sex, and a
scale that grows after age 40. Three correlated markers come along for
later vignettes.

```{r data}
ref <- norm_simulate(400, kind = "gaussian", scale = "age", seed = 1)
ref$participant_id <- paste0("R", seq_len(nrow(ref)))
head(ref[, c("participant_id", "age", "sex", "site", "y")])
```

## Specify and fit

A `norm_spec()` names a distribution family and one formula per
distributional parameter. Covariates are whatever the formulas mention;
other columns are ignored. The Gaussian family with a constant scale fits
an ordinary additive model; a non-constant scale formula switches to a
location-scale model.

```{r fit}
spec <- norm_spec(
  family = norm_gaussian(),
  location = ~ s(age, k = 8) + sex,
  scale = ~ s(age, k = 5)
)
fit <- norm_fit(spec, data = ref, outcomes = "y", id = participant_id)
fit
```

The centile chart is the fitted conditional distribution, not a scatter
of the training rows. Held-out people are overlaid as points; the bands
are the 5th-95th and 25th-75th centile ranges.

```{r centiles}
val <- norm_simulate(150, kind = "gaussian", scale = "age", seed = 5)
autoplot(fit, type = "centiles", by = sex, newdata = val)
```

## Which distribution? A model ladder scored out of sample

Whether a Gaussian is enough, or whether the scale, skewness, or tail
weight need to vary with age, is an empirical question that in-sample
fit cannot settle. `norm_select()` takes a list of candidate
specifications ordered from simplest to most flexible, cross-fits each,
and applies a lexicographic rule on the out-of-fold scores: drop
uncalibrated candidates, rank the rest by log score with CRPS as the
tie-breaker, keep everything within one paired standard error of the
best, and choose the simplest survivor. Covariate-dependent skew or tail
must earn a minimum log-score gain before it is accepted.

```{r ladder}
ladder <- list(
  gaussian_const = norm_spec(norm_gaussian(), location = ~ s(age, k = 8) + sex),
  gaussian_scale = spec,
  shash_scale = norm_spec(norm_shash(), location = ~ s(age, k = 8) + sex,
                          scale = ~ s(age, k = 5))
)
sel <- norm_select(ladder, data = ref, outcomes = "y", folds = 4)
sel$comparison[, c("model", "mean_log_score", "crps", "var_z", "cover_95",
                   "calibrated", "selected")]
```

Read the table from the left: every candidate passed the calibration
gate, and the log scores differ by a few hundredths of a nat per
observation. The two Gaussian levels are separated by roughly one
paired standard error and SHASH, whose skew and tail parameters are
constant here, is admitted without a shape penalty; with 400 rows and
four folds the selection among calibrated candidates is close and a
different seed can move it. What the ladder guarantees is that an
uncalibrated level is never chosen and that extra flexibility must show
an out-of-sample gain. `sel$selected` is the winning spec and
`sel$crossfit` its out-of-fold scores, which are the right input for the
joint model in the multi-site vignette.

## Check calibration on held-out people

`norm_assess()` needs data the model has not seen. It reports proper
scores (`mean_log_score`, `crps`), the PCNtoolkit-style point metrics
(`smse`, `ev`, `cor`), marginal calibration of the normal scores, and
tail rates at four levels.

```{r assess}
assessment <- norm_assess(fit, newdata = val)
assessment$overall[, c("mean_log_score", "standardized_log_score", "crps", "ev", "cor")]
assessment$marginal[, c("n", "mean_z", "var_z", "cover_50", "cover_80", "cover_95")]
```

A calibrated model has `mean_z` near 0, `var_z` near 1, and observed
coverage on the diagonal. The worm plot shows where the predictive shape
is wrong: a tilt means the scale is off, a U or inverted U means skew,
an S means the tails.

```{r calibration}
autoplot(assessment, type = "calibration")
autoplot(assessment, type = "worm")
```

## Place new observations

`predict()` returns one row per person and outcome. `uncertainty =
"total"` (the default) integrates the score over the uncertainty of the
fitted curves; `"conditional"` uses the point estimate and is faster for
diagnostics. `support` says whether the person's covariates are inside
the reference data (`"in"`), near its boundary (`"edge"`), outside it
(`"out"`, where `z` is masked unless `allow_extrapolation = TRUE`), or in
a factor level the reference never saw (`"new_group"`).

```{r scores}
scores <- predict(fit, newdata = val, type = "scores")
scores[1:5, c(".id", "observed", "median", "centile", "z", "tail_prob", "support")]
```

The scores are derived from a full predictive distribution, which you can
ask for directly. `predict(type = "distribution")` returns a
[distributional](https://pkg.mitchelloharawild.com/distributional/)
vector per outcome, so `cdf()`, `quantile()`, `hilo()`, `mean()`,
`variance()`, and `generate()` all apply, and `as_scores()` turns a
distribution and an observation into the score columns above.

```{r distributions}
dists <- predict(fit, newdata = val[1:3, ], type = "distribution",
                 uncertainty = "conditional")
dists$y
hilo(dists$y, 90)
as_scores(dists$y, val$y[1:3])
```

`tidy()` summarises the fit per outcome, `glance()` per fit, and
`augment()` returns the target rows with `.z_<outcome>` and
`.centile_<outcome>` columns, one row per person.

```{r tidiers}
tidy(fit)[, c("outcome", "family", "status", "n", "edf")]
glance(fit)
augment(fit, val[1:3, ], uncertainty = "conditional")[, c("age", "sex", "y", ".z_y", ".centile_y", ".support")]
```

## Threshold exceedances are not abnormalities

`norm_flag()` marks `|z|` above a threshold and adjusts the two-sided tail
probabilities for multiple comparisons. It also reports how many
exceedances a calibrated model would produce by chance, which is the
number to compare against before reading anything into the flagged rows.

```{r flag}
flagged <- norm_flag(scores, threshold = 2)
attr(flagged, "observed_exceedances")
attr(flagged, "expected_exceedances")
flagged[flagged$exceedance, c(".id", "observed", "centile", "z", "tail_prob", "fdr")]
```

A person can sit in the tail because the chart is right and they are
unusual, because their covariates are outside the reference support, or
because the model is mis-specified. The score table keeps those apart:
`z` represents a centile, `support` says whether that centile can be
trusted, and the assessment says whether the model deserves it. For
multi-site data, site transport, and joint deviation across many
outcomes, continue with `vignette("brain-charts-across-sites")`; for
repeated visits, `vignette("longitudinal-change")`.
