---
title: "Getting started with referent"
description: "Fit, score, and interpret one distributional reference model in a complete first workflow."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with referent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message = FALSE, warning = FALSE}
library(referent)
```

You have a reference cohort, an outcome such as a regional brain volume, and
covariates that explain ordinary population variation. You want to place new
observations in the distribution expected for people with those covariates.
This page takes one default path from those inputs to an interpretable score.

A reference model estimates the full conditional distribution, not just a
mean curve. A centile is the distribution's CDF at the observed value; `z` is
`qnorm(centile)`; and `tail_prob` is the smaller tail. None of them is a
clinical label.

# What data do you need?

The reference rows define the population. Target rows are the observations you
will score. Here both are deterministic simulations with an age trend, sex
effect, and age-dependent spread.

```{r data}
reference <- ref_simulate(400, kind = "gaussian", scale = "age", seed = 1)
target <- ref_simulate(150, kind = "gaussian", scale = "age", seed = 5)
head(reference[, c("age", "sex", "site", "y")])
```

In your data, use stable participant identifiers and fit only on rows that
belong to the declared reference population. `referent` never decides who
belongs in that population for you.

# How do you fit and score the first model?

`ref_spec()` declares one formula for each distributional parameter.
`ref_fit()` fits that specification, and `predict()` places held-out rows in
the fitted distribution.

```{r first-fit}
spec <- ref_spec(
  family = ref_gaussian(),
  location = ~ s(age, k = 8) + sex,
  scale = ~ s(age, k = 5)
)
fit <- ref_fit(spec, data = reference, outcomes = "y")
target <- target[ref_support(fit, target)$support == "in", ]
scores <- predict(fit, newdata = target, type = "scores")
scores[1:5, c(".id", "observed", "median", "centile", "z", "tail_prob", "support")]
```

```{r first-fit-checks, include = FALSE}
stopifnot(
  nrow(scores) == nrow(target) && nrow(scores) > 100,
  all(is.finite(scores$centile)),
  all(scores$centile > 0 & scores$centile < 1),
  stats::sd(scores$z) > 0.5
)
```

Read one row as a chain: `observed` is the measurement, `median` is the
reference model's central prediction, `centile` is its conditional rank, and
`z` expresses that rank on a standard-normal scale. `support` says whether the
covariates justify interpolation.

# What does the fitted chart show?

```{r centiles, fig.alt="Fitted median and central centile bands across age, split by sex, with held-out target observations overlaid."}
autoplot(fit, type = "centiles", by = sex, newdata = target)
```

Notice that the bands widen with age because the scale formula varies with
age. The points are held out: the figure shows how new observations sit in the
reference distribution, not how closely the model redraws its training data.
It is a calibration aid, not evidence that the reference cohort is clinically
appropriate.

# When should you trust a score?

Support is part of the result, not an optional diagnostic. By default,
`referent` masks probability scores outside the reference covariate range, for
missing predictors, and for unseen factor levels.

```{r support}
edge_cases <- target[1:3, ]
edge_cases$age <- c(median(reference$age), max(reference$age) + 20, NA_real_)
ref_support(fit, edge_cases)[, c("support", "d2")]
predict(fit, edge_cases)[, c(".id", "z", "support", "status")]
```

Do not remove the mask just to obtain a number. First decide whether the row is
a data error, needs a better reference cohort, or belongs to a documented
transport analysis. The [troubleshooting article](/articles/troubleshooting/)
shows those recovery paths.

# How do you use the predictive distribution directly?

`predict(type = "distribution")` returns a `distributional` vector per
outcome. The same object supplies centiles, intervals, simulation, and scores,
so those summaries cannot silently drift apart.

```{r distributions}
dists <- predict(fit, target[1:3, ], type = "distribution",
                 uncertainty = "conditional")$y
hilo(dists, 90)
as_scores(dists, target$y[1:3])[, c("centile", "z", "tail_prob")]
```

# Does the model deserve interpretation?

Assess on rows that were not used for fitting. Here `uncertainty = "total"`
matches the default score call above and includes uncertainty in the fitted
curves.

```{r assess}
assessment <- ref_assess(fit, target, uncertainty = "total")
assessment$marginal[, c("n", "mean_z", "var_z", "cover_50", "cover_95")]
assessment$overall[, c("mean_log_score", "crps", "rmse")]
```

```{r assess-checks, include = FALSE}
m <- assessment$marginal
stopifnot(
  all(is.finite(m$mean_z)),
  all(is.finite(m$var_z) & m$var_z > 0.25),
  all(m$cover_95 > 0.80 & m$cover_95 <= 1)
)
```

A calibrated model has `mean_z` near 0, `var_z` near 1, and observed coverage
near its nominal level. Those are sampling targets, not exact pass/fail values
for one small data set. Use the validation workflow next when you need model
selection, conditional diagnostics, and tail checks.

# What should you report about large deviations?

```{r flags}
flags <- ref_flag(scores, threshold = 2)
c(
  observed = attr(flags, "observed_exceedances"),
  expected = attr(flags, "expected_exceedances")
)
```

A large `|z|` can mean a genuinely unusual observation, unsupported
covariates, or a miscalibrated model. Report the reference definition,
held-out calibration, support status, and expected exceedance count before
interpreting any flagged row.

Continue with [Validate and choose a reference model](/articles/validate-reference/).
For site transport, go to [Brain charts across sites](/articles/brain-charts-across-sites/).
