---
title: "Getting started with referent"
description: "Fit, score, and interpret one distributional reference model in a complete first workflow."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with referent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message = FALSE, warning = FALSE}
library(referent)
```

You have a reference cohort, an outcome such as a regional brain volume, and
covariates that explain ordinary population variation. You want to place new
observations in the distribution expected for people with those covariates.
This page takes one default path from those inputs to an interpretable score.

A reference model estimates the full conditional distribution, not just a
mean curve. A centile is the distribution's CDF at the observed value; `z` is
`qnorm(centile)`; and `tail_prob` is the smaller tail. None of them is a
clinical label.

# FAQ: choosing covariates and a reference population

These decisions are made before the first `ref_fit()` call and no function
in the package makes them for you. The answers below are the defaults a
careful analyst would reach for; each links to the place where the package
gives you the evidence to revise them.

**Which covariates?** Age and sex are the usual conditioning set for a brain
chart: they explain ordinary population variation you do not want to call a
deviation. Put age in a smooth, `s(age, k = 8)`, and sex as a factor. Anything
you add becomes part of the question the score answers: conditioning on
education, for example, turns "is this volume unusual for a 60-year-old
woman?" into "... for a 60-year-old woman with this much education", which may
or may not be what you want to know.

**Site as a random effect, not a covariate.** Scanners shift measurements, but
a new participant may come from a scanner that contributed no reference rows.
`s(site, bs = "re")` estimates a shrunk level per site and returns the
population curve for an unseen level (reported as `support == "new_group"`),
while a fixed `site` factor has no prediction at all for a level it never saw.
[Brain charts across sites](https://bbuchsbaum.github.io/referent/articles/brain-charts-across-sites/)
covers the transport choices that follow.

**Head size: covariate or offset?** A covariate `log(icv)` lets the data decide
how volume scales with intracranial volume; the score then asks whether the
volume is unusual *for that head size*. An offset, `offset(log(icv))` under
`transform = "log"`, fixes the scaling at proportional and scores the volume
*as a fraction of* head size. The two differ when the scaling is not
proportional (it usually is not, across regions), and they answer different
questions: prefer the covariate unless you have a reason to assert
proportionality.

```{r icv}
proportions <- ref_simulate(400, seed = 11)
proportions$icv <- exp(rnorm(nrow(proportions), 14, 0.1))
proportions$volume <- exp(0.8 * log(proportions$icv) + 0.02 * proportions$age +
                            rnorm(nrow(proportions), 0, 0.05))
as_covariate <- ref_spec(ref_gaussian(), ~ s(age, k = 6) + sex + log(icv),
                         transform = "log")
as_offset <- ref_spec(ref_gaussian(), ~ s(age, k = 6) + sex + offset(log(icv)),
                      transform = "log")
rbind(
  covariate = tidy(ref_fit(as_covariate, proportions, "volume"))[, c("status", "edf")],
  offset = tidy(ref_fit(as_offset, proportions, "volume"))[, c("status", "edf")]
)
```

Both fit; the offset model spends one parameter fewer because it never
estimates the slope on `log(icv)`, and it is misspecified here because the
true exponent is 0.8, not 1.

**Who belongs in the reference?** Everyone whose distribution you want to call
"expected". A reference of healthy controls gives scores relative to health; a
population-based reference includes the prevalence of disease in that
population. The package never decides this because no covariate column can
encode the decision: write the inclusion rule down, apply it before
`ref_fit()`, and record it in `ref_freeze(criteria = )` so it travels with
the reference.

**How many local controls does `ref_adapt()` need?** Adaptation estimates a
location offset and, optionally, a log-scale offset, each shrunk toward zero
by a ridge worth `location_prior_n` and `scale_prior_n` pseudo-observations.
The location offset is a shrunk mean residual: with 30 controls its standard
error is about `1 / sqrt(30)`, or 0.18 in `z` units, and it is usable with
far fewer. The scale offset needs several times more, because the variance
of a variance estimate is larger. The
[site-transport comparison](https://bbuchsbaum.github.io/referent/articles/brain-charts-across-sites/)
sets out that arithmetic for 30 controls, and its conclusion is the general
one: adapt location first, adapt scale only when local n justifies it, and
calibrate by site only with still more.

**What does `|z| > 2` mean?** Under a calibrated model, 4.55% of rows from
the reference population exceed it by chance, so the count you observe is
only informative against that expectation. `ref_flag()` reports it as the
`expected_exceedances` attribute alongside the observed count and an
FDR-adjusted tail probability per row. A flagged row is an exceedance with a
known chance rate, not a diagnosis.

**Gaussian or SHASH?** Start Gaussian. Switch to `ref_shash()` when held-out
scores are skewed or heavy-tailed (`skew_z` and `excess_kurtosis_z` in
`ref_assess()$marginal`), and let `ref_select()` arbitrate: its calibration
gate removes candidates whose out-of-fold coverage and tail rates are wrong,
and its one-standard-error rule keeps the simplest model that predicts as
well as the best. A SHASH fit can also let skew and tail weight vary with a
covariate. Use that flexibility sparingly; it is estimated from the tails of
the data, where there is least information.

```{r shash-shape}
skewed <- ref_simulate(400, kind = "shash", skew = 0.6, tail = 1, seed = 12)
shape_spec <- ref_spec(
  family = ref_shash(),
  location = ~ s(age, k = 6) + sex,
  scale = ~ s(age, k = 5),
  skew = ~ s(age, k = 4),
  tail = ~ 1
)
shape_fit <- ref_fit(shape_spec, skewed, outcomes = "y")
shape_fit
```

Read further in
[Validate and choose a reference model](https://bbuchsbaum.github.io/referent/articles/validate-reference/)
(`vignette("validate-reference", package = "referent")`).

# What data do you need?

The reference rows define the population. Target rows are the observations you
will score. Here both are deterministic simulations with an age trend, sex
effect, and age-dependent spread.

```{r data}
reference <- ref_simulate(400, kind = "gaussian", scale = "age", seed = 1)
target <- ref_simulate(150, kind = "gaussian", scale = "age", seed = 5)
head(reference[, c("age", "sex", "site", "y")])
```

In your data, use stable participant identifiers and fit only on rows that
belong to the declared reference population. `referent` never decides who
belongs in that population for you.

# How do you fit and score the first model?

`ref_spec()` declares one formula for each distributional parameter.
`ref_fit()` fits that specification, and `predict()` places held-out rows in
the fitted distribution.

```{r first-fit}
spec <- ref_spec(
  family = ref_gaussian(),
  location = ~ s(age, k = 8) + sex,
  scale = ~ s(age, k = 5)
)
fit <- ref_fit(spec, data = reference, outcomes = "y")
target <- target[ref_support(fit, target)$support == "in", ]
scores <- predict(fit, newdata = target, type = "scores")
scores[1:5, c(".id", "observed", "median", "centile", "z", "tail_prob", "support")]
```

```{r first-fit-checks, include = FALSE}
stopifnot(
  nrow(scores) == nrow(target) && nrow(scores) > 100,
  all(is.finite(scores$centile)),
  all(scores$centile > 0 & scores$centile < 1),
  stats::sd(scores$z) > 0.5
)
```

Read one row as a chain: `observed` is the measurement, `median` is the
reference model's central prediction, `centile` is its conditional rank, and
`z` expresses that rank on a standard-normal scale. `support` says whether the
covariates justify interpolation.

# What does the fitted chart show?

```{r centiles, fig.alt="Fitted median and central centile bands across age, split by sex, with held-out target observations overlaid."}
autoplot(fit, type = "centiles", by = sex, newdata = target)
```

Notice that the bands widen with age because the scale formula varies with
age. The points are held out: the figure shows how new observations sit in the
reference distribution, not how closely the model redraws its training data.
It is a calibration aid, not evidence that the reference cohort is clinically
appropriate.

# When should you trust a score?

Support is part of the result, not an optional diagnostic. By default,
`referent` masks probability scores outside the reference covariate range, for
missing predictors, and for unseen factor levels.

```{r support}
edge_cases <- target[1:3, ]
edge_cases$age <- c(median(reference$age), max(reference$age) + 20, NA_real_)
ref_support(fit, edge_cases)[, c("support", "d2")]
predict(fit, edge_cases)[, c(".id", "z", "support", "status")]
```

Do not remove the mask just to obtain a number. First decide whether the row is
a data error, needs a better reference cohort, or belongs to a documented
transport analysis. The
[troubleshooting article](https://bbuchsbaum.github.io/referent/articles/troubleshooting/)
(`vignette("troubleshooting", package = "referent")`) shows those recovery
paths.

# How do you use the predictive distribution directly?

`predict(type = "distribution")` returns a `distributional` vector per
outcome. The same object supplies centiles, intervals, simulation, and scores,
so those summaries cannot silently drift apart.

```{r distributions}
dists <- predict(fit, target[1:3, ], type = "distribution",
                 uncertainty = "conditional")$y
hilo(dists, 90)
as_scores(dists, target$y[1:3])[, c("centile", "z", "tail_prob")]
```

`generate()` draws synthetic observations from the same distributions, which
is how you simulate a reference-like cohort with given covariates, for a
power calculation or to check a downstream pipeline on data with no real
participants in it.

```{r generate}
set.seed(1)
synthetic <- generate(dists, times = 4)
synthetic
```

# Does the model deserve interpretation?

Assess on rows that were not used for fitting. Here `uncertainty = "total"`
matches the default score call above and includes uncertainty in the fitted
curves.

```{r assess}
assessment <- ref_assess(fit, target, uncertainty = "total")
assessment$marginal[, c("n", "mean_z", "var_z", "cover_50", "cover_95")]
assessment$overall[, c("mean_log_score", "crps", "rmse")]
```

```{r assess-checks, include = FALSE}
m <- assessment$marginal
stopifnot(
  all(is.finite(m$mean_z)),
  all(is.finite(m$var_z) & m$var_z > 0.25),
  all(m$cover_95 > 0.80 & m$cover_95 <= 1)
)
```

A calibrated model has `mean_z` near 0, `var_z` near 1, and observed coverage
near its nominal level. Those are sampling targets, not exact pass/fail values
for one small data set. Use the validation workflow next when you need model
selection, conditional diagnostics, and tail checks.

# What should you report about large deviations?

```{r flags}
flags <- ref_flag(scores, threshold = 2)
c(
  observed = attr(flags, "observed_exceedances"),
  expected = attr(flags, "expected_exceedances")
)
```

A large `|z|` can mean a genuinely unusual observation, unsupported
covariates, or a miscalibrated model. Report the reference definition,
held-out calibration, support status, and expected exceedance count before
interpreting any flagged row.

Continue with [Validate and choose a reference model](https://bbuchsbaum.github.io/referent/articles/validate-reference/).
For site transport, go to [Brain charts across sites](https://bbuchsbaum.github.io/referent/articles/brain-charts-across-sites/).
