---
title: "Synthetic growth reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Synthetic growth reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.8
)
```

A height-like growth measure with age and sex. No brain data. The walkthrough
is the usual referent path: fit a reference, check that the predictive
distribution is calibrated, place new people in it, and treat large deviations
as threshold exceedances rather than abnormalities.

```{r setup}
library(referent)
set.seed(1)
```

## Fit a reference

```{r fit}
ref <- norm_simulate(400, kind = "gaussian", scale = "age", seed = 1)
ref$participant_id <- paste0("R", seq_len(nrow(ref)))
spec <- norm_spec(
  family = norm_gaussian(),
  location = ~ s(age, k = 8) + sex,
  scale = ~ s(age, k = 5)
)
fit <- norm_fit(spec, data = ref, outcomes = "y", id = participant_id)
fit
```

The chart is the fitted conditional distribution, not the scatter of the
training rows. Overlay held-out people only as points. Facet by sex so each
panel has its own location–scale curve.

```{r centiles}
val <- norm_simulate(150, kind = "gaussian", scale = "age", seed = 2)
val$participant_id <- paste0("V", seq_len(nrow(val)))
autoplot(fit, type = "centiles", by = sex, newdata = val)
```

Visit counts and the age mix by site are composition diagnostics, not part of
the fitted geometry.

```{r composition}
autoplot(fit, type = "visits", data = ref, id = participant_id)
autoplot(fit, type = "coverage", data = ref, time = age, by = site)
```

## Check calibration on held-out people

```{r assess}
assessment <- norm_assess(fit, newdata = val)
assessment$overall
assessment$marginal[, c(".outcome", "mean_z", "var_z", "cover_50", "cover_80", "cover_95")]
```

Coverage should sit near the nominal bands. A worm plot that hugs zero means
the PIT is close to uniform after mapping to normal scores.

```{r calibration}
autoplot(assessment, type = "calibration")
```

```{r worm}
autoplot(assessment, type = "worm")
```

## Place new observations

`predict()` returns one row per person–outcome. The subject id declared at
fit time is copied onto `.id` when that column is present in `newdata`.

```{r scores}
scores <- predict(fit, newdata = val, type = "scores", uncertainty = "conditional")
scores[1:5, c(".id", "observed", "median", "centile", "z", "tail_prob", "support")]
```

The scores are derived from a full predictive distribution, which you can
ask for directly. `predict(type = "distribution")` returns a tibble with
`.id` and one `<distribution>` column per outcome; these are
[distributional](https://pkg.mitchelloharawild.com/distributional/) vectors,
so `cdf()`, `quantile()`, `hilo()`, `mean()`, `variance()`, and
`generate()` all apply, and `as_scores()` turns a distribution and an
observation into the score columns above.

```{r distributions}
dists <- predict(fit, newdata = val[1:3, ], type = "distribution",
                 uncertainty = "conditional")
dists
hilo(dists$y, 90)
as_scores(dists$y, val$y[1:3])
```

`augment()` returns the same information in wide form, one row per person
with `.z_<outcome>` and `.centile_<outcome>` columns, and `tidy()` gives a
one-row-per-outcome summary of the fit.

```{r augment}
augment(fit, val[1:3, ], uncertainty = "conditional")
tidy(fit)
```

The score table is an ordinary data frame; subset it the way you would any
other (`dplyr::filter()` keeps the class).

```{r filter}
large <- scores[abs(scores$z) > 2 & !is.na(scores$z), ]
nrow(large)
large[, c(".id", "observed", "centile", "z", "tail_prob")]
```

`flag()` marks threshold exceedances and reports how many you should expect
from a well-calibrated Gaussian tail. It does not label anyone abnormal.

```{r flag}
flagged <- flag(scores, threshold = 2)
attr(flagged, "observed_exceedances")
attr(flagged, "expected_exceedances")
```

## What a large deviation is not

A person can sit in the tail because the reference chart is right, because the
covariates are outside support, or because the model is mis-specified. The
score table keeps those apart: `z` is a representation of a centile, and
`support` says whether that centile is even trustworthy. Use
`flag()` and ordinary subsetting to inspect tails. Do not treat either as a
diagnosis.
