---
title: "Longitudinal change"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal change}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  dpi = 96
)
```

A cross-sectional chart is a trajectory of distributions, not a
distribution of trajectories. It says what the population looks like at
each age; it says nothing about how one person moves from one visit to
the next. People track centiles, and whether a follow-up value is
surprising depends on how strongly their earlier values predict it.

`referent` treats this as a dynamic extension of the reference model.
The marginal fit maps each visit to a normal score
`Z = qnorm(F(y | x))`; a Gaussian process on those scores describes
within-person dependence across time; everything longitudinal is then a
conditional distribution under that process.

```{r setup}
library(referent)
```

## Longitudinal reference data

Subjects have two to four visits at irregular intervals. The normal
score of `y` is generated as a stable subject component plus a
Matern-3/2 process in time plus measurement noise, with variance
fractions 0.5, 0.35, and 0.15.

```{r data}
long <- norm_simulate(1200, kind = "longitudinal", seed = 1)
truth <- attr(long, "truth")
c(stable = truth$tau_b^2, dynamic = truth$tau_g^2, measurement = truth$sigma_e^2,
  ell = truth$ell)
head(long[, c("participant_id", "visit", "age", "sex", "y")])
```

## The marginal reference, then the process

The marginal model is the usual cross-sectional fit. `norm_dynamics()`
then estimates the process on the normal scores. By default it uses
out-of-fold scores from a subject-level cross-fit, so that in-sample
shrinkage does not masquerade as within-person dependence.

```{r fit}
spec <- norm_spec(norm_gaussian(), location = ~ s(age, k = 6) + sex, scale = ~1)
fit <- norm_fit(spec, data = long, outcomes = "y", id = participant_id)
dyn <- norm_dynamics(fit, data = long, id = participant_id, time = age)
dyn
```

The print shows the estimated variance fractions of the stable, dynamic,
and measurement components, the correlation at the median lag with a
standard error, and the Matern length-scale. The `components` table
holds the same numbers.

```{r components}
dyn$components
```

The kernel plot is the implied correlation between two normal scores of
the same person as a function of the time between them. The jump at lag
zero is the measurement nugget; the plateau is the stable rank.

```{r kernel}
autoplot(dyn, type = "kernel")
```

## Identifiability and the stable model

Three components cannot be separated from two-visit data at one fixed
interval. With a single lag, any dependence structure collapses to one
number, the correlation at that lag, and the Matern length-scale is not
identified. `norm_dynamics()` detects this and reduces to the stable
model (`norm_process("stable")`): a subject intercept plus nugget, one
parameter. Everything downstream still works; it just cannot
extrapolate to other intervals, and forecasts at other lags are flagged
`extrapolated_lag`.

```{r stable}
two_visit <- long[long$visit <= 2, ]
two_visit <- two_visit[ave(two_visit$visit, two_visit$participant_id, FUN = length) == 2, ]
two_visit$age <- ave(two_visit$age, two_visit$participant_id, FUN = function(a) a[1] + 2 * (seq_along(a) - 1))
dyn_fixed <- norm_dynamics(fit, data = two_visit, id = participant_id, time = age, crossfit = 0)
dyn_fixed$components[, c("process", "identified", "ell_identified", "median_lag", "r_median_lag")]
```

Separating measurement noise from rank dynamics needs either short
repeat intervals or three or more visits per subject;
`dyn$identifiability$measurement` records whether that was possible.
When too few subjects have repeats, the process is reported as not
identified and every history-conditioned quantity downstream is `NA`.

## Six longitudinal quantities

The design separates six questions that are often conflated.

| Quantity | Question | Where |
|---|---|---|
| Chart velocity | How does a reference centile itself move with age? | `norm_derivative()` |
| Observed velocity | At what raw rate did this person change? | `norm_transition()$observed_velocity` |
| Expected velocity | What rate was expected given this person's history? | `norm_transition()$expected_velocity` |
| Velocity centile | Where does the observed rate fall in the conditional distribution? | `norm_transition()$velocity_centile` |
| Innovation Z | How surprising is the follow-up after conditioning on history? | `norm_transition()$innovation_z` |
| Change Z | How unusual is the change in the unconditional distribution of changes? | `norm_transition()$change_z` |

Chart velocity is a property of the population curve, not of any person:

```{r derivative}
grid <- data.frame(age = c(30, 50, 70), sex = factor("F", levels = c("F", "M")))
norm_derivative(fit, grid, with_respect_to = age, centiles = c(0.1, 0.5, 0.9))
```

The others come from `norm_transition()`, one row per consecutive pair
of visits. Innovation Z is the history-conditioned score
`(Z_2 - m) / s`, where `m` and `s` are the conditional mean and SD of
the new score given all earlier ones; with one prior visit it reduces
to the Z-gain `(Z_2 - r Z_1) / sqrt(1 - r^2)`. Change Z is the
difference `Z_2 - Z_1` standardised by its own SD `sqrt(2(1 - r))`.
Both are N(0, 1) under the reference process; the naive difference
`Z_2 - Z_1` is not.

```{r transition}
held <- norm_simulate(2400, kind = "longitudinal", seed = 7)
tr <- norm_transition(dyn, data = held, id = participant_id, time = age)
tr[1:4, c(".id", ".dt", "observed_velocity", "expected_velocity",
          "velocity_centile", "innovation_z", "change_z", "support")]
```

On held-out subjects from the same generator both scores are calibrated,
and the rate of `|z| > 1.96` is near 5%. The naive difference is not.

```{r calib}
naive <- with(tr, qnorm(end_centile) - qnorm(start_centile))
rbind(
  innovation = c(var = var(tr$innovation_z, na.rm = TRUE),
                 fpr = mean(abs(tr$innovation_z) > 1.96, na.rm = TRUE)),
  change = c(var = var(tr$change_z, na.rm = TRUE),
             fpr = mean(abs(tr$change_z) > 1.96, na.rm = TRUE)),
  naive = c(var = var(naive, na.rm = TRUE),
            fpr = mean(abs(naive) > 1.96, na.rm = TRUE))
)
autoplot(dyn, type = "calibration", data = held, id = participant_id, time = age)
```

```{r innovation}
autoplot(tr, type = "innovation")
```

## Forecasts

`norm_forecast()` conditions the marginal distribution at future times
on a subject's history and returns one `dist_conditioned()` predictive
distribution per time. The fan starts at the last observed visit; its
width at short horizons reflects the measurement nugget and grows toward
the marginal spread as the dynamic component decorrelates. Times beyond
the reference lag range are reported as `extrapolated_lag`.

```{r forecast}
subject <- held[held$participant_id == 3, ]
fc <- norm_forecast(dyn, history = subject, times = max(subject$age) + c(0.5, 1, 2, 3, 5))
fc$summary
autoplot(fc, type = "fan")
```

The forecast distribution obeys every generic of the package: a later
observation can be scored against it with `as_scores()`, which gives
the same innovation Z that `norm_transition()` would.

```{r score-forecast}
as_scores(fc$dist[2], 10.5)[, c("centile", "z", "tail_prob")]
```

## Reading the output

* `components`: fractions of normal-score variance that are stable,
  dynamic, and measurement. A large measurement fraction means visits
  are noisy and short-interval change is mostly noise; a large stable
  fraction means people hold their rank and innovation Z is sensitive.
* `r_median_lag` with its SE: the one number the data identify most
  directly. If the SE is large, do not interpret `ell`.
* `support` on transitions and forecasts: `in`, `extrapolated_lag`,
  `unsupported_age`, `insufficient_history`, or `unidentified`. Only
  `in` rows have a calibrated innovation Z.
* `calibrated` on transitions is `TRUE` only when measurement noise was
  separable; otherwise the innovation Z still conditions on history but
  treats all short-lag dependence as rank dynamics.
