---
title: "Coming from PCNtoolkit"
description: "Translate PCNtoolkit concepts and metrics into referent without overstating equivalence."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coming from PCNtoolkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

PCNtoolkit users will recognise most of `referent`. The estimand is the
same: a conditional predictive distribution of an outcome given
covariates, estimated on a reference population and used to place new
people. The differences are in what is exposed. `referent` returns the
distribution itself as a first-class object and derives every score from
its CDF, it separates site *adaptation* (shifting the distribution) from
*calibration* (re-mapping its probabilities), and it does not define an
abnormality index.

```{r setup}
library(referent)
ref <- ref_simulate(300, seed = 1)
val <- ref_simulate(150, seed = 2)
fit <- ref_fit(ref_spec(ref_gaussian(), ~ s(age, k = 6) + sex), ref, "y")
scores <- predict(fit, newdata = val)
assessment <- ref_assess(fit, newdata = val)
```

# Concepts

| PCNtoolkit | referent | Notes |
|---|---|---|
| normative model (`BLR`, `HBR`) | [`ref_spec()`](https://bbuchsbaum.github.io/referent/reference/ref_spec/) + [`ref_fit()`](https://bbuchsbaum.github.io/referent/reference/ref_fit/) | one `mgcv` model per outcome; Gaussian or SHASH families with formulas per parameter; GPR belongs to the legacy PCNtoolkit interface, not 1.x |
| response variable | `outcomes =` | several outcomes fit as a panel, kept wide |
| covariates, batch effects | formula terms | `s(site, bs = "re")` plays the role of a batch random effect |
| `predict()` / `estimate()` | `predict(fit, newdata)` | returns scores or `<distribution>` columns |
| cross-validation (`cvfolds`) | `ref_crossfit()` | out-of-fold scores for every reference row; the deployment refit is attached |
| warp (`SHASHb`, `SinArcsinh`) | `ref_shash()` | sinh-arcsinh likelihood via `mgcv::shash()` |
| `NormData` `batch_effects` | `s(site, bs = "re")`; `ref_calibrate(by = )` | the batch variable is a random intercept in the formula; group-specific probability re-mapping is a separate step |
| HBR random slopes | not available; `s(age, site, bs = "fs")` | `bs = "re"` is a random intercept only; `mgcv`'s factor-smooth basis gives each site its own age curve shrunk toward the shared one (see below) |
| HBR Beta likelihood | not available | families are Gaussian and SHASH on a monotone transform of the outcome; a bounded outcome can be logit-transformed by hand before fitting |
| `GPR` | not available | smooth terms (`s(age)`, `te(age, x)`) take the place of a GP mean; their uncertainty enters `uncertainty = "total"` through the coefficient covariance |
| warped BLR (`WarpSinArcsinh`, `WarpBoxCox`) | `ref_spec(transform = )` and `ref_shash()` | a Box-Cox or log transform of the response is `transform =`; a sinh-arcsinh likelihood is `ref_shash()`; the two combine |
| `fit_predict` | `ref_crossfit()` + `ref_fit()` | cross-fit for out-of-fold reference scores, then fit on all rows for deployment |
| site transfer / `adapt` / `transfer` | `ref_adapt()` | shrunk location (and scale) offsets on the family's own density, trajectory frozen |
| (no direct analogue) | `ref_calibrate()` | monotone re-mapping of PIT values; optional `by = site` |
| `extend` | refit on pooled data | `ref_fit(spec, rbind(reference, new_site))` is the extend operation; the reference changes for every site |
| `merge` / federated models | none | the unit of sharing is a frozen bundle (`ref_freeze()`), not model parameters to be merged |
| harmonised outputs (`Y_harmonized`) | `predict(type = "harmonised")` | conditional quantile mapping out of the site effect (or to a chosen `to` level); the deviation score itself is already site-free |
| extreme value / abnormality | not provided | see below |

## Harmonised values

When a downstream analysis needs raw values on one scale, the conditional
quantile map gives them: a value at the 80th centile of its own site's
predictive becomes the 80th centile of the target predictive.

```{r harmonise}
sites <- ref_simulate(300, sites = 3, site_shift = c(0, 1, -1), seed = 5)
site_fit <- ref_fit(
  ref_spec(ref_gaussian(), ~ s(age, k = 5) + sex + s(site, bs = "re")),
  sites, "y"
)
harmonised <- predict(site_fit, sites, type = "harmonised")
rbind(raw = tapply(sites$y, sites$site, mean),
      harmonised = tapply(harmonised$y, harmonised$site, mean))
```

## Random slopes by site

`s(site, bs = "re")` shifts the level per site. If sites also differ in the
shape of the age trend, `mgcv`'s factor-smooth interaction,
`s(age, site, bs = "fs")`, fits one smooth per site with a shared penalty
that shrinks each curve toward the common one. It is accepted anywhere a
formula is, and an unseen site still receives the population curve with
`support == "new_group"`.

```{r fs}
multi <- ref_simulate(600, sites = 5, seed = 4)
fs_spec <- ref_spec(
  ref_gaussian(),
  location = ~ s(age, k = 6) + sex + s(age, site, bs = "fs", k = 5)
)
fs_fit <- ref_fit(fs_spec, multi, "y")
tidy(fs_fit)[, c("status", "edf_s(age)", "edf_s(age,site)")]
```

```{r fs-check, include = FALSE}
stopifnot(tidy(fs_fit)$status == "ok")
```

This is not HBR: the site curves are penalised deviations from a shared
smooth, not draws from a hierarchical prior with its own estimated
hyperparameters, and `ref_adapt()` transports only the level and scale, not
the slope. Use it when the reference contains enough rows per site to
support a curve each, and cross-fit with `strata = site` to check that it
earns its effective degrees of freedom.

# Metrics

`ref_assess()` reports related evaluation quantities alongside proper scoring
rules, but shared names do not guarantee the same estimand. The table labels
each relationship rather than silently treating near-matches as identical.
The PCNtoolkit side follows its
[1.x evaluation-metrics definitions](https://pcntoolkit.readthedocs.io/en/stable/pages/tutorials/13_evaluation_metrics.html),
not the legacy 0.35 interface.

| PCNtoolkit | referent column | Equivalence | Relationship |
|---|---|---|---|
| [`Z`](https://pcntoolkit.readthedocs.io/en/stable/pages/tutorials/13_evaluation_metrics.html) | `z` | Conceptual analogue | Both place an observation in a predictive distribution. `referent` always computes `qnorm(centile)` from the distribution CDF, including non-Gaussian families. |
| `yhat` | `median` | Exact when `yhat` is the predictive median | For a Gaussian the median also equals the mean. Check the model family's definition before comparing other families. |
| legacy BLR `s2` | `variance(dist)` | Model-specific analogue | Both can include observation and coefficient uncertainty, but PCNtoolkit 1.x does not expose `s2` as a general output and HBR posterior averaging need not match `referent`'s mixture distribution. |
| `MSLL` | no exact column | Not numerically equivalent | PCNtoolkit subtracts a Gaussian baseline fitted to the sample being scored and uses lower-is-better loss. `standardized_log_score` uses the frozen reference-sample baseline and higher-is-better log score. Rescore both against one declared baseline before comparing. |
| `SMSE` | `smse` | Near-equivalent | Both divide MSE by outcome variance, but PCNtoolkit uses a population variance denominator while R's `var()` uses a sample denominator. On the same complete rows, referent's value is PCNtoolkit's value times `(n - 1) / n`. |
| `EV` (`EXPV`) | `ev` | Algebraically equivalent on nonconstant complete data | Both compute `1 - var(residual) / var(y)`. Constant outcomes and filtering of non-finite rows can produce different edge behaviour. |
| `Rho` | no exact column | Different statistic | PCNtoolkit 1.x reports Spearman rank correlation. `cor` is Pearson correlation of the observed value and predictive median. |
| `RMSE` | `rmse` | Exact on the same included rows | Missing and non-finite row handling must also match. |
| `MACE` | `mace` | Exact only without batch effects | Both use the mean absolute calibration gap on a centile grid. With batch combinations, PCNtoolkit 1.3.0 averages the batch-level values equally; `referent` currently pools rows. |
| `ShapiroW` | `shapiro_w` | Same statistic | Shapiro-Wilk W of Z; `referent` subsamples 5000 scores with a fixed seed when n is larger, because `shapiro.test()` accepts no more. |
| `Skew`, `Kurtosis` | `skew_z`, `excess_kurtosis_z` | Related, not exact | PCNtoolkit 1.3.0 uses SciPy's bias-corrected skewness and excess kurtosis. `referent` reports unadjusted standardised moments (and also uses excess kurtosis). The difference is material for small samples. |
| (none) | `mean_log_score`, `crps` | referent additions | Proper scores of the full predictive distribution; larger log score and smaller CRPS are better. |

One implementation detail matters for external rescoring: with outcome
standardisation enabled, PCNtoolkit 1.3.0 maps predictions and centiles back to
the outcome scale but returns `logp` on the standardised scale. Subtract
`log(sd(y_train))` (using the population standard deviation used by the
scaler), plus any response-transform Jacobian, before comparing response-scale
log scores.

```{r metrics}
assessment$overall[, c("standardized_log_score", "smse", "ev", "cor", "rmse", "crps")]
```

If you need PCNtoolkit's `Rho`, compute it explicitly from the held-out score
rows. Pearson and Spearman answer different questions and can disagree even on
the same predictions.

```{r rho}
metric_rows <- assessment$scores
c(
  pearson = cor(metric_rows$observed, metric_rows$median, method = "pearson"),
  spearman_rho = cor(metric_rows$observed, metric_rows$median, method = "spearman")
)
```

```{r rho-check, include = FALSE}
rho_values <- c(
  pearson = cor(metric_rows$observed, metric_rows$median, method = "pearson"),
  spearman_rho = cor(metric_rows$observed, metric_rows$median, method = "spearman")
)
stopifnot(all(is.finite(rho_values)), abs(diff(rho_values)) > 1e-4)
```

The normality metrics are in the marginal table. The chunk after it
recomputes them from the same held-out scores (`assessment$scores`; the
`scores` object above was computed under `uncertainty = "total"` and differs
slightly), which is also how to get them on a different centile grid or
without the kurtosis offset.

```{r normality}
assessment$marginal[, c("mean_z", "var_z", "skew_z", "excess_kurtosis_z",
                        "shapiro_w", "mace")]
held <- assessment$scores
z <- held$z[is.finite(held$z)]
u <- held$centile[is.finite(held$centile)]
grid <- c(0.05, 0.25, 0.5, 0.75, 0.95)
c(
  MACE = mean(vapply(grid, function(q) abs(q - mean(u <= q)), numeric(1))),
  ShapiroW = unname(shapiro.test(z)$statistic),
  Skew = mean((z - mean(z))^3) / sd(z)^3,
  Kurtosis = mean((z - mean(z))^4) / sd(z)^4
)
```

```{r normality-check, include = FALSE}
m <- assessment$marginal
stopifnot(
  abs(mean((z - mean(z))^3) / sd(z)^3 - m$skew_z) < 0.05,
  abs(mean((z - mean(z))^4) / sd(z)^4 - 3 - m$excess_kurtosis_z) < 0.05,
  abs(unname(shapiro.test(z)$statistic) - m$shapiro_w) < 1e-6
)
```

```{r z}
d <- predict(fit, newdata = val[1:3, ], type = "distribution")$y
data.frame(
  yhat = median(d),
  s2 = variance(d),
  Z = scores$z[1:3]
)
```

`Z` here includes uncertainty in the fitted curves (`uncertainty = "total"`,
the default). That is closest to legacy BLR's coefficient-aware predictive
variance, but it is not a general claim of numerical equivalence across
PCNtoolkit model families. In particular, PCNtoolkit HBR averages
draw-specific Z scores and quantiles, while `referent` derives them from a
predictive mixture. Those are different estimands even with identical draws.
`uncertainty = "conditional"` gives the plug-in score.

## What numerical agreement means

The executable comparison target is PCNtoolkit 1.3.0. Fixed Gaussian
distributions and transformed-distribution algebra are checked as exact;
SHASH distributions are checked after an explicit parameter conversion;
deliberately matched fitted models use prediction-level tolerances. When the
estimators differ, agreement is assessed by held-out proper scores and
calibration rather than by forcing their parameters or predictions to match.

These are independent claims: better held-out performance cannot excuse a
failure of shared distribution algebra, and numerical agreement does not by
itself establish superiority. The complete classifications, tolerances, and
CI tiers are recorded in the repository's
`docs/design/pcntoolkit-validation-contract.md`, with current machine-readable
results under `docs/evidence/pcntoolkit/v1.3.0/`. Both paths are linked from
the repository README.

# Site adaptation

PCNtoolkit's adaptation re-estimates site-specific parameters of a
pre-trained model from a small local sample. `ref_adapt()` does the
same with the shared trajectory frozen, estimating a location offset
(and optionally a log-scale offset) per site by penalised maximum
likelihood, with shrinkage set in pseudo-observations.

```{r adapt}
local <- ref_simulate(40, site_shift = 1, seed = 3)
adapted <- ref_adapt(fit, data = local, parameters = c("location", "scale"))
adapted$adaptation
```

PCNtoolkit has no separate calibration step. `referent` keeps the two
ideas apart because they fix different problems: adaptation moves the
distribution when a new site's level or spread differs; calibration
re-maps probabilities when the distributional family is wrong in a way
that is common to everyone. `vignette("brain-charts-across-sites")`
compares both against a refit on held-out data.

# What is deliberately missing

PCNtoolkit offers extreme-value and abnormality statistics that summarise
how far a profile lies in the tails. `referent` does not provide an
abnormality index. A large `|z|` is a threshold exceedance whose expected
count under calibration is known; `ref_flag()` reports that count and an
FDR-adjusted tail probability, and `ref_joint()` gives a calibrated
joint centile across outcomes. Whether an exceedance means anything is a
question for the study design, not the scoring function.

To state the rest plainly, `referent` does not do the following.

* **Hierarchical Bayesian regression.** There is no MCMC, no posterior over
  site slopes, and no Beta or other bounded likelihood. Site effects are
  penalised random intercepts (or factor smooths); uncertainty in the curves
  is Gaussian in the coefficients.
* **Gaussian-process regression.** Smooth terms play that role.
* **Harmonisation as a separate model.** `predict(type = "harmonised")`
  maps each value through its own site's predictive CDF and the target
  site's quantile function, so it removes exactly the site effect the
  reference model estimated and nothing else. There is no ComBat-style
  empirical-Bayes step, and the deviation score, which is site-free by
  construction, is the quantity meant for downstream use.
* **Model merging or federated fitting.** A reference is fitted on data in one
  place and shipped as a bundle. Extending a reference means refitting on the
  pooled data.
* **Classification or clinical thresholds.** No function maps a score to a
  label.

Continue with [Troubleshooting reference-model workflows](https://bbuchsbaum.github.io/referent/articles/troubleshooting/)
(`vignette("troubleshooting", package = "referent")`),
or return to [Getting started with referent](https://bbuchsbaum.github.io/referent/articles/getting-started/) for the
default end-to-end workflow.
