---
title: "Coming from PCNtoolkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coming from PCNtoolkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

PCNtoolkit users will recognise most of `referent`. The estimand is the
same: a conditional predictive distribution of an outcome given
covariates, estimated on a reference population and used to place new
people. The differences are in what is exposed. `referent` returns the
distribution itself as a first-class object and derives every score from
its CDF, it separates site *adaptation* (shifting the distribution) from
*calibration* (re-mapping its probabilities), and it does not define an
abnormality index.

```{r setup}
library(referent)
ref <- ref_simulate(300, seed = 1)
val <- ref_simulate(150, seed = 2)
fit <- ref_fit(ref_spec(ref_gaussian(), ~ s(age, k = 6) + sex), ref, "y")
scores <- predict(fit, newdata = val)
assessment <- ref_assess(fit, newdata = val)
```

## Concepts

| PCNtoolkit | referent | Notes |
|---|---|---|
| normative model (`BLR`, `HBR`, `GPR`) | `ref_spec()` + `ref_fit()` | one `mgcv` model per outcome; Gaussian or SHASH families with formulas per parameter |
| response variable | `outcomes =` | several outcomes fit as a panel, kept wide |
| covariates, batch effects | formula terms | `s(site, bs = "re")` plays the role of a batch random effect |
| `predict()` / `estimate()` | `predict(fit, newdata)` | returns scores or `<distribution>` columns |
| cross-validation (`cvfolds`) | `ref_crossfit()` | out-of-fold scores for every reference row; the deployment refit is attached |
| warp (`SHASHb`, `SinArcsinh`) | `ref_shash()` | sinh-arcsinh likelihood via `mgcv::shash()` |
| site transfer / `adapt` | `ref_adapt()` | shrunk location (and scale) offsets on the family's own density, trajectory frozen |
| (no direct analogue) | `ref_calibrate()` | monotone re-mapping of PIT values; optional `by = site` |
| extreme value / abnormality | not provided | see below |

## Metrics

`ref_assess()` reports the PCNtoolkit evaluation metrics under their
usual definitions, alongside proper scoring rules.

| PCNtoolkit | referent column | Relationship |
|---|---|---|
| `Z` | `z` | `qnorm(centile)`, evaluated from the smaller tail in log space so very large deviations stay finite |
| `yhat` | `median` | the predictive median; for a Gaussian this equals the mean |
| `s2` | `variance(dist)` | predictive variance of the distribution returned by `predict(type = "distribution")` |
| `MSLL` | `-standardized_log_score` | mean log score minus that of an unconditional Gaussian; sign flipped so larger is better |
| `SMSE` | `smse` | mean squared residual divided by the variance of the observed outcome |
| `EV` (`EXPV`) | `ev` | explained variance, `1 - var(residual) / var(y)` |
| `Rho` | `cor` | Pearson correlation of observed and predicted median |
| `RMSE` | `rmse` | root mean squared residual |
| (none) | `mean_log_score`, `crps` | proper scores of the full predictive distribution |

```{r metrics}
assessment$overall[, c("standardized_log_score", "smse", "ev", "cor", "rmse", "crps")]
```

```{r z}
d <- predict(fit, newdata = val[1:3, ], type = "distribution")$y
data.frame(
  yhat = median(d),
  s2 = variance(d),
  Z = scores$z[1:3]
)
```

`Z` here includes the uncertainty of the fitted curves
(`uncertainty = "total"`, the default), which is what PCNtoolkit's
predictive variance also does. `uncertainty = "conditional"` gives the
plug-in score.

## Site adaptation

PCNtoolkit's adaptation re-estimates site-specific parameters of a
pre-trained model from a small local sample. `ref_adapt()` does the
same with the shared trajectory frozen, estimating a location offset
(and optionally a log-scale offset) per site by penalised maximum
likelihood, with shrinkage set in pseudo-observations.

```{r adapt}
local <- ref_simulate(40, site_shift = 1, seed = 3)
adapted <- ref_adapt(fit, data = local, parameters = c("location", "scale"))
adapted$adaptation
```

PCNtoolkit has no separate calibration step. `referent` keeps the two
ideas apart because they fix different problems: adaptation moves the
distribution when a new site's level or spread differs; calibration
re-maps probabilities when the distributional family is wrong in a way
that is common to everyone. `vignette("brain-charts-across-sites")`
compares both against a refit on held-out data.

## What is deliberately missing

PCNtoolkit offers extreme-value and abnormality statistics that summarise
how far a profile lies in the tails. `referent` does not provide an
abnormality index. A large `|z|` is a threshold exceedance whose expected
count under calibration is known; `ref_flag()` reports that count and an
FDR-adjusted tail probability, and `ref_joint()` gives a calibrated
joint centile across outcomes. Whether an exceedance means anything is a
question for the study design, not the scoring function.
