---
title: "Troubleshooting reference-model workflows"
description: "Diagnose missing scores, transport boundaries, partial fits, weak local corrections, and unidentified dynamics."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting reference-model workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This article starts where [Getting started with referent](https://bbuchsbaum.github.io/referent/articles/getting-started/)
(`vignette("getting-started", package = "referent")`)
ends. You have a
fitted reference model, but a score is missing, a transported site behaves
unexpectedly, or a longitudinal extension refuses to make a claim. Each
section follows the same path: recognize the symptom, identify the cause,
and make the smallest defensible fix.

```{r fit-reference}
library(referent)

reference <- ref_simulate(240, seed = 10)
held_out <- ref_simulate(100, seed = 11)
spec <- ref_spec(ref_gaussian(), location = ~ s(age, k = 5) + sex)
fit <- ref_fit(spec, data = reference, outcomes = "y")
```

# Why is a score missing for an ordinary-looking row?

## Symptom

A used covariate is missing, or the row lies outside the range represented by
the reference data. `referent` keeps the fitted median where it can, but masks
the probability-scale quantities that would otherwise look more trustworthy
than they are.

```{r missing-covariate}
in_rows <- which(ref_support(fit, held_out)$support == "in")[1:4]
incomplete <- held_out[in_rows, ]
incomplete$age[2] <- NA_real_

missing_scores <- predict(fit, incomplete, uncertainty = "conditional")
missing_scores[, c(".row", "z", "support", "status")]
```

```{r check-missing-covariate, include = FALSE}
stopifnot(
  is.na(missing_scores$z[2]),
  missing_scores$support[2] == "unknown",
  missing_scores$status[2] == "missing_predictor",
  all(is.finite(missing_scores$z[-2]))
)
```

An out-of-range covariate is different from a missing one, but it receives the
same conservative treatment on the score scale.

```{r outside-support}
outside <- held_out[in_rows[1:3], ]
outside$age[1] <- max(reference$age) + 20

outside_scores <- predict(fit, outside, uncertainty = "conditional")
outside_scores[, c(".row", "median", "z", "support")]
```

```{r check-outside-support, include = FALSE}
stopifnot(
  outside_scores$support[1] == "out",
  is.finite(outside_scores$median[1]),
  is.na(outside_scores$z[1])
)
```

## Cause

The formulas in `ref_spec()` define the covariates the model actually needs.
An `NA` in one of those columns gives `status == "missing_predictor"` and
`support == "unknown"`. A finite value beyond the observed reference range
gives `support == "out"`. Neither condition is repaired by imputing or
silently extending the fitted curve.

## Fix

Supply the missing predictor from a defensible data source, or restrict the
analysis to the declared reference domain. Use `ref_support(fit, newdata)`
before a large scoring run to audit support without fitting or scoring again.
Do not use `allow_extrapolation = TRUE` merely to turn an unsupported age into
a number.

# Why is a transported site's score still marked `new_group`?

## Symptom

The shared model was fitted on sites A-D. Site E has local reference controls,
and `ref_adapt()` estimates a site-E offset, but the original support record
still says that E was absent from the training reference.

```{r fit-site-reference}
site_spec <- ref_spec(
  ref_gaussian(),
  location = ~ s(age, k = 5) + sex + s(site, bs = "re")
)
site_fit <- ref_fit(site_spec, reference, outcomes = "y")

site_e <- ref_simulate(
  600, sites = 5,
  site_shift = c(0, 0, 0, 0, 1),
  seed = 12
)
site_e <- site_e[site_e$site == "E", ]
local_controls <- site_e[1:20, ]
site_e_target <- site_e[21:50, ]
adapted <- ref_adapt(site_fit, local_controls, by = site)
```

By default, unseen-group probability columns remain masked. After fitting an
explicit transport model from local controls, you can deliberately request the
transported scores while retaining `support == "new_group"` as provenance.

```{r score-transported-site}
masked <- predict(adapted, site_e_target, uncertainty = "conditional")
transported <- predict(
  adapted,
  site_e_target,
  uncertainty = "conditional",
  allow_extrapolation = TRUE
)

data.frame(
  default_z = head(masked$z, 3),
  transported_z = head(transported$z, 3),
  support = head(transported$support, 3)
)
```

```{r check-transported-site, include = FALSE}
stopifnot(
  all(is.na(masked$z)),
  all(is.finite(transported$z)),
  all(transported$support == "new_group")
)
```

## Cause

Adaptation changes the predictive location or scale; it does not rewrite the
historical fact that site E was outside the original site's factor levels.
`allow_extrapolation = TRUE` exposes the adapted distribution, while the
support column preserves that boundary.

## Fix

Use this override only after fitting a deliberate transport step from an
appropriate local reference sample. Keep `support` in every exported score
table, report how many local controls estimated the transport, and validate on
different local observations. See
[Brain charts across sites](https://bbuchsbaum.github.io/referent/articles/brain-charts-across-sites/) for the full
comparison among population scoring, adaptation, calibration, and refitting.

# Why does a small local correction look weak or fail to calibrate?

## Symptom

An adaptation offset is smaller than the raw local mean difference, or a
calibration object built from a tiny sample leaves `calibrated == FALSE`.

```{r inspect-small-adaptation}
site_e_offset <- adapted$adaptation$offsets$y$E
data.frame(
  n = site_e_offset$n,
  location = site_e_offset$location,
  location_se = site_e_offset$location_se
)
```

A single probability value cannot define the package's monotone calibration
map, so no row is reported as calibrated.

```{r tiny-calibration}
tiny_calibration <- ref_calibrate(
  site_fit,
  data = local_controls[1, , drop = FALSE],
  by = site,
  uncertainty = "conditional"
)
tiny_scores <- predict(
  tiny_calibration,
  site_e_target,
  uncertainty = "conditional",
  allow_extrapolation = TRUE
)

data.frame(
  calibration_n = tiny_calibration$calibration$n,
  any_mapped = any(tiny_scores$calibrated)
)
```

```{r check-small-local-sets, include = FALSE}
stopifnot(
  site_e_offset$n == nrow(local_controls),
  is.finite(site_e_offset$location),
  is.finite(site_e_offset$location_se),
  !any(tiny_scores$calibrated)
)
```

## Cause

`ref_adapt()` deliberately shrinks local offsets toward the shared reference;
the amount depends on the local sample size and the prior strength.
`ref_calibrate()` instead estimates an empirical probability map. With fewer
than two finite calibration values there is no map, and with only a modest
number its tail resolution is necessarily coarse.

## Fix

Choose the correction that matches the failure. Adapt when a scanner or domain
plausibly shifts location or scale while preserving the shared trajectory.
Calibrate when held-out diagnostics show probability miscalibration that a
location/scale shift cannot repair. Reserve separate local rows for evaluation,
report the local sample size, and inspect centre, spread, coverage, and tails;
do not judge a correction on the rows that estimated it.

# Why did one outcome fail while the rest of the panel fitted?

## Symptom

`ref_fit()` returns a panel object, but `tidy()` reports
`unsupported_type` for one outcome.

```{r failed-outcome}
panel <- reference
panel$group_label <- factor(ifelse(panel$y > median(panel$y), "high", "low"))

panel_fit <- ref_fit(
  spec,
  data = panel,
  outcomes = c("y", "group_label")
)
tidy(panel_fit)[, c("outcome", "status", "message")]
```

```{r check-failed-outcome, include = FALSE}
panel_status <- tidy(panel_fit)
stopifnot(
  panel_status$status[panel_status$outcome == "y"] == "ok",
  panel_status$status[panel_status$outcome == "group_label"] == "unsupported_type"
)
```

## Cause

The current model families define continuous predictive distributions for
numeric outcomes. A factor, character, or logical outcome is recorded as a
failed panel member instead of aborting every other outcome.

## Fix

Select the intended numeric measurements explicitly. Convert an encoded
numeric column only when its scale has a defensible quantitative meaning; do
not turn categories into arbitrary integer codes to make the status disappear.
Inspect `tidy(fit)` before deployment and decide whether any partial panel is
acceptable for the study.

# Why is an assessment too optimistic?

## Symptom

The model looks well calibrated when assessed on the same rows used to fit it.
`ref_assess()` now rejects this reuse by default. An explicitly requested
training diagnostic remains labelled `in_sample`; it is not validation.

```{r compare-assessment-sources}
training_assessment <- ref_assess(fit, newdata = reference, allow_in_sample = TRUE)
held_out_assessment <- ref_assess(fit, newdata = held_out)

data.frame(
  sample = c("training", "held out"),
  in_sample = c(training_assessment$in_sample, held_out_assessment$in_sample),
  var_z = c(training_assessment$marginal$var_z, held_out_assessment$marginal$var_z),
  cover_95 = c(training_assessment$marginal$cover_95,
               held_out_assessment$marginal$cover_95)
)
```

```{r check-assessment-sources, include = FALSE}
stopifnot(
  isTRUE(training_assessment$in_sample),
  !isTRUE(held_out_assessment$in_sample),
  all(is.finite(held_out_assessment$marginal$var_z)),
  all(is.finite(held_out_assessment$marginal$cover_95))
)
```

## Cause

The fitted rows influenced the curve against which they are being scored.
Smoothness selection and parameter estimation can make their deviations look
smaller than deviations for genuinely new observations.

## Fix

Pass a held-out reference sample to `ref_assess()`. When a separate sample is
not available, use out-of-fold scores from `ref_crossfit()` and keep whole
subjects together with `cluster =` when rows repeat within person.

```{r crossfit-reference}
oof <- ref_crossfit(spec, reference, outcomes = "y", folds = 3)
table(oof$.in_sample)
```

```{r check-crossfit-reference, include = FALSE}
stopifnot(nrow(oof) == nrow(reference), !any(oof$.in_sample))
```

# Why are longitudinal change scores all `NA`?

## Symptom

The marginal model fits, but `ref_dynamics()` reports an unidentified process
and `ref_transition()` returns no innovation or change Z-scores.

```{r sparse-longitudinal-design}
sparse <- ref_simulate(80, seed = 20)
sparse$participant_id <- seq_len(nrow(sparse))
sparse$participant_id[1:3] <- sparse$participant_id[4:6]

sparse_fit <- ref_fit(
  spec,
  data = sparse,
  outcomes = "y",
  id = participant_id
)
sparse_dynamics <- ref_dynamics(
  sparse_fit,
  data = sparse,
  id = participant_id,
  time = age,
  crossfit = 0
)
sparse_dynamics$components[, c("identified", "stable", "dynamic", "measurement")]
```

```{r unidentified-transitions}
sparse_transitions <- ref_transition(
  sparse_dynamics,
  data = sparse,
  id = participant_id,
  time = age
)
sparse_transitions[, c(".id", "innovation_z", "change_z", "support")]
```

```{r check-unidentified-dynamics, include = FALSE}
stopifnot(
  !sparse_dynamics$processes$y$identified,
  nrow(sparse_transitions) == 3,
  all(is.na(sparse_transitions$innovation_z)),
  all(is.na(sparse_transitions$change_z)),
  all(sparse_transitions$support == "unidentified")
)
```

## Cause

Only three subjects have repeat observations here. That is not enough to
identify within-person dependence, so the package fails closed instead of
turning starting values into longitudinal evidence. Even when change is
identifiable, separating stable rank, dynamic variation, and measurement noise
requires informative visit timing, such as three or more visits for some
subjects or short-interval repeats.

## Fix

Treat visit design as part of the estimand. Collect repeated observations from
enough subjects, include varied lags when estimating a time-decaying process,
and inspect `components`, `identified`, `ell_identified`, and support before
interpreting change. [Longitudinal change](https://bbuchsbaum.github.io/referent/articles/longitudinal-change/) explains
the stable fixed-lag fallback, innovation Z, change Z, velocity, and
forecasting.

# Where should you go next?

Return to [Getting started with referent](https://bbuchsbaum.github.io/referent/articles/getting-started/) for the ordinary
fit-score-assess path. Use
[Brain charts across sites](https://bbuchsbaum.github.io/referent/articles/brain-charts-across-sites/) for site transport,
joint scores, and reporting; use
[Longitudinal change](https://bbuchsbaum.github.io/referent/articles/longitudinal-change/) for repeated visits; and use
[Coming from PCNtoolkit](https://bbuchsbaum.github.io/referent/articles/pcntoolkit/) when translating an existing
PCNtoolkit workflow.
The function reference for `predict.ref_fit()`, `ref_support()`,
`ref_assess()`, and `ref_dynamics()` gives the exact status and return-value
contracts used here.
