---
title: "Validate and choose a reference model"
description: "Choose distributional flexibility out of sample and diagnose marginal, conditional, and tail calibration."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validate and choose a reference model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message = FALSE, warning = FALSE}
library(referent)
reference <- ref_simulate(400, kind = "gaussian", scale = "age", seed = 1)
validation <- ref_simulate(300, kind = "gaussian", scale = "age", seed = 8)
```

The first model that converges is not necessarily a trustworthy reference.
You need to decide how much distributional flexibility the data support and
then test the chosen model on people who did not fit it. This page keeps those
two jobs separate: cross-fitting chooses a candidate within the reference
cohort; a held-out validation cohort checks the final fit.

# Which distributional model earns its complexity?

Order candidates from the simplest plausible model to more flexible ones.
Every `ref_select()` criterion is computed from out-of-fold predictions, so a
candidate cannot win by explaining the same rows that fitted it.

```{r ladder}
ladder <- list(
  gaussian_constant = ref_spec(
    ref_gaussian(), location = ~ s(age, k = 8) + sex
  ),
  gaussian_varying_scale = ref_spec(
    ref_gaussian(), location = ~ s(age, k = 8) + sex,
    scale = ~ s(age, k = 5)
  ),
  shash_varying_scale = ref_spec(
    ref_shash(), location = ~ s(age, k = 8) + sex,
    scale = ~ s(age, k = 5)
  )
)
selection <- ref_select(ladder, reference, outcomes = "y", folds = 4)
selection$comparison[, c(
  "model", "mean_log_score", "crps", "mean_z", "var_z",
  "cover_95", "calibrated", "selected"
)]
```

```{r ladder-checks, include = FALSE}
tab <- selection$comparison
finite_oof_z <- selection$crossfit$z[is.finite(selection$crossfit$z)]
stopifnot(
  sum(tab$selected) == 1,
  all(is.finite(tab$mean_log_score[tab$status %in% c("ok", "partial")])),
  length(finite_oof_z) > 0.8 * nrow(selection$crossfit),
  stats::sd(finite_oof_z) > 0.5
)
```

Read the table as a sequence of gates. Failed or unstable fits leave first.
Calibration is checked before log score and CRPS. The paired one-standard-error
rule then keeps candidates whose predictive performance is not distinguishable
from the best, and the simplest survivor wins. A selected model is the most
defensible member of this ladder, not proof that the ladder contained truth.
Fold-specific support checks can mask a few boundary rows when one fold's
training range is narrower than the full reference range; the comparison
reports the finite out-of-fold scoring set rather than filling those values.

# How do you validate the selected model honestly?

Fit the selected specification to all reference rows, then assess a separate
cohort. Declare the uncertainty regime explicitly so the reported diagnostics
can be reproduced from `predict()`.

```{r held-out}
fit <- ref_fit(selection$selected, reference, outcomes = "y")
assessment <- ref_assess(fit, validation, uncertainty = "total")
assessment$overall[, c(
  "mean_log_score", "standardized_log_score", "crps", "smse", "ev", "rmse"
)]
assessment$marginal[, c(
  "n", "mean_z", "var_z", "cover_50", "cover_80", "cover_95", "cover_99"
)]
```

```{r held-out-checks, include = FALSE}
marginal <- assessment$marginal
stopifnot(
  !assessment$in_sample,
  all(is.finite(marginal$mean_z)),
  all(is.finite(marginal$var_z) & marginal$var_z > 0),
  all(is.finite(assessment$overall$mean_log_score))
)
```

`mean_z` and `var_z` diagnose location and spread on the normal-score scale.
Coverage checks several central intervals. Log score and CRPS assess the full
distribution rather than only its median. Compare models on the same held-out
rows and uncertainty regime.

# Where is the distribution wrong?

```{r calibration-plot, fig.alt="Observed coverage versus nominal coverage for the held-out validation cohort."}
autoplot(assessment, type = "calibration")
```

Points near the diagonal indicate marginal coverage. Departures identify
which interval widths miss their nominal rates, but they do not reveal whether
the error changes with age, sex, or site.

```{r worm-plot, fig.alt="Worm plot of held-out normal-score quantile departures from the standard normal reference."}
autoplot(assessment, type = "worm")
```

A tilted worm suggests a scale error; curvature suggests skew or tail
misfit. Treat that shape as a diagnosis to test, not automatic permission to
add a more flexible family. The QQ plot shows the same ordered scores
against the standard normal without the detrending, which makes gross tail
departures easier to see and small ones harder.

```{r qq-plot, fig.alt="Quantile-quantile plot of held-out normal scores against the standard normal."}
autoplot(assessment, type = "qq")
```

The `tail` table quantifies what the plot shows: observed versus expected
exceedance rates at several tail levels, with standard errors.

```{r tail}
assessment$tail
```

# Does calibration drift across covariates or groups?

The conditional table checks every covariate used by the fit. Numeric
covariates receive light diagnostic smooths; factors receive one row per
level. The p-values test no drift, while the drift columns describe its size.

```{r conditional}
assessment$conditional[, c(
  ".outcome", "covariate", "level", "n",
  "location_drift", "location_p", "scale_drift", "scale_p"
)]
```

If you ask for `by = site`, the marginal table is also split by site. Do this
only when each group has enough held-out rows to estimate coverage and tails.

```{r by-site}
by_site <- ref_assess(fit, validation, by = site, uncertainty = "total")
by_site$marginal[, c(".group", "n", "mean_z", "var_z", "cover_95")]
```

# What should happen after a failed check?

Do not tune and validate on the same rows. Revise the candidate ladder or
covariates using reference-only cross-fitting, then return once to an untouched
validation cohort. If failure is concentrated at a new scanner, use a
transport analysis instead of bending the shared trajectory to that scanner.
If scores are missing, diagnose support before model complexity.

Continue with [Brain charts across sites](https://bbuchsbaum.github.io/referent/articles/brain-charts-across-sites/)
(`vignette("brain-charts-across-sites", package = "referent")`), or
use [Troubleshooting reference models](https://bbuchsbaum.github.io/referent/articles/troubleshooting/) when a fit or
score does not behave as expected.
